---
title: "Bayesian methods and modern statistics"
subtitle: "Fall 2024"
mainfont: Lato
format:
   html: 
    page-layout: full
    toc: false
---

# Schedule

```{r load-libraries, echo = F, message = F, eval = F}
library(tidyverse)
library(knitr)
```

```{r smart-date-setup, echo = F, message = F, eval = F}
# set these:
firstDay = as.Date("2024-08-26")
firstDayID = "Mon"
lastDay = as.Date("2024-12-06")
lectureDays = c("Mon", "Wed")
labDays = c("Thu")
holidays = as.Date(c("2024-09-02", "2024-10-14", "2024-10-15", "2024-11-27", "2024-11-28", "2024-11-29"))

### render table
allDates = seq(from = firstDay, to = lastDay, by = 'day')
schedule = data.frame(allDates,
                        dayOfWeek = weekdays(allDates, abbreviate=T)) %>%
  filter(dayOfWeek %in% c(lectureDays, labDays))
schedule = schedule %>%
  mutate(lab = ifelse(dayOfWeek %in% labDays, TRUE, FALSE)) %>%
  mutate(Topic = ifelse(allDates %in% holidays, "NO CLASS", "")) %>%
  mutate(Date = format(allDates, format = "%b %d")) %>%
  mutate(Date = paste(dayOfWeek, Date)) %>%
  mutate(Week = rep("", nrow(schedule))) %>%
  mutate(Prepare = rep("", nrow(schedule))) %>%
  mutate(AE = rep("", nrow(schedule))) %>%
  mutate(Assignment = rep("", nrow(schedule))) %>%
  mutate(Exam = rep("", nrow(schedule))) %>%
  mutate(Project = rep("", nrow(schedule))) %>%
  select(Week, Date, Topic, Prepare, AE, Assignment, Exam, Project)
# to be pasted into console then pasted into markdown:
schedule %>%
  kable(format = "markdown")
```

```{=html}
<style>
    .heatMap {
        width: 100%;
        text-align: center;
    }
    .heatMap th {
        word-wrap: break-word;
        text-align: center;
    }
<!--     .heatMap tr:nth-of-type(2n+1) { -->
<!--   background: #FFF; -->
<!-- } -->
    .heatMap tr:nth-of-type(3n) { background: #D3D3D3; }
</style>
```
::: heatMap
| Week | Date       | Topic                                               | Reading   |                                            Notes                                             | Assignment                 |
|:----------|:----------|:----------|-----------|:--------------:|:----------|
| 1    | Mon Aug 26 | intro, history, notation                            | Ch. 2     |                                                                                              | [hw 0](/hw/hw00.html)      |
|      | Wed Aug 28 | probability, exchangeability                        | Ch. 2     |       [🗒](/notes/lec01-probability.html) [📝](/notes/Probability-Exchangeability.pdf)        | [hw 1](/hw/hw01.html)      |
|      | Thu Aug 29 | lab: welcome                                        |           |                               [💻](/slides/lab0-welcome.html)                                | [hello R](/labs/lab0.html) |
| 2    | Mon Sep 02 | NO CLASS                                            | Ch. 3     |                                                                                              |                            |
|      | Wed Sep 04 | beta-binomial model                                 | Ch. 3     |               [🗒](/notes/lec02-estimation.html) [📝](/notes/beta-binomial.pdf)               | [hw 2](/hw/hw02.html)      |
|      | Thu Sep 05 | lab: MLE and MAP estimator                          |           |                                   [💻](/slides/lab1.html)                                    |                            |
| 3    | Mon Sep 09 | Poisson-gamma model, exp families                   | Ch. 3     | [🗒](notes/lec03-Poisson-gamma-exp-families.html) [📝](/notes/Poisson-gamma-exp-families.pdf) |                            |
|      | Wed Sep 11 | reliability (conf. intervals, hpd, Laplace approx.) | Ch. 3     |         [🗒](/notes/lec04-reliability.html) [📝](/notes/reliability-handwritten.pdf)          | [hw 3](/hw/hw03.html)      |
|      | Thu Sep 12 | lab: exp. families and transformations              |           |                             [💻](/slides/lab-exp-families.html)                              |                            |
| 4    | Mon Sep 16 | intro to Monte Carlo                                | Ch. 4     |         [🗒](/notes/lec05-introMonteCarlo.html) [📝](/notes/posteriorPredictive.pdf)          |                            |
|      | Wed Sep 18 | predictive checks and MC error                      | Ch. 4     |                       [🗒](/notes/lec06-MonteCarloPredictionError.html)                       | [hw 4](/hw/hw04.html)      |
|      | Thu Sep 19 | lab: mixture densities                              |           |                               [💻](/slides/lab-mixtures.html)                                |                            |
| 5    | Mon Sep 23 | the normal model                                    | Ch. 5     |             [🗒](/notes/lec07-normalModel.html)[📝](/notes/completeTheSquare.pdf)             |                            |
|      | Wed Sep 25 | the normal model II                                 | Ch. 5     |                                 [📝](notes/normalNotes2.pdf)                                 |                            |
|      | Thu Sep 26 | lab: normal data                                    |           |                                [💻](/slides/lab-normal.html)                                 |                            |
| 6    | Mon Sep 30 | review                                              |           |                                                                                              |                            |
|      | Wed Oct 02 | [Exam I](/notes/exam-notes.html)                    |           |                                                                                              |                            |
|      | Thu Oct 03 | NO LAB                                              |           |                                                                                              |                            |
| 7    | Mon Oct 07 | Metropolis algorithm                                | Ch. 10    |                         [🗒](/notes/lec08-Metropolis-algorithm.html)                          |                            |
|      | Wed Oct 09 | MCMC diagnostics                                    | Ch. 6     |                                 [🗒](/notes/lec09-MCMC.html)                                  | [hw 5](/hw/hw05.html)      |
|      | Thu Oct 10 | lab: Metropolis algo.                               |           |                               [💻](/slides/lab9-MH-MCMC.html)                                |                            |
| 8    | Mon Oct 14 | NO CLASS                                            |           |                                                                                              |                            |
|      | Wed Oct 16 | wrap-up diagnostics and MH                          | Ch. 6, 10 |                             [📝](notes/ergodicMarkovChains.pdf)                              |                            |
|      | Thu Oct 17 | lab: MCMC and conf. bands                           |           |                               [💻](/slides/lab-conf-band.html)                               |                            |
| 9    | Mon Oct 21 | Gibbs sampling                                      | Ch. 6     |                             [🗒](notes/lec10-gibbs-sampling.html)                             |                            |
|      | Wed Oct 23 | Multivariate normal                                 | Ch. 7     |                                  [🗒](notes/lec11-mvn.html)                                   | [hw 6](/hw/hw06.html)      |
|      | Thu Oct 24 | lab: MCMC diagnostics                               |           |                            [💻](slides/lab6-mcmc-d-practice.html)                            |                            |
| 10   | Mon Oct 28 | intro to Bayesian regression                        | Ch. 9     |            [🗒](notes/lec12-regression-intro.html)[📝](notes/intro-regression.pdf)            |                            |
|      | Wed Oct 30 | Bayesian regression II                              | Ch. 9     |                          [🗒](notes/lec13-BayesianRegression2.html)                           | [hw 7](/hw/hw07.html)      |
|      | Thu Oct 31 | lab: rstanarm                                       |           |                                 [💻](slides/lab8-rstan.html)                                 |                            |
| 11   | Mon Nov 04 | Hierarchical modeling                               |           |                                                                                              | hw 8                       |
|      | Wed Nov 06 | Estimators                                          |           |                                                                                              |                            |
|      | Thu Nov 07 | lab: estimators                                     |           |                                                                                              |                            |
| 12   | Mon Nov 11 | review                                              |           |                                                                                              |                            |
|      | Wed Nov 13 | Exam II                                             |           |                                                                                              |                            |
|      | Thu Nov 14 | lab                                                 |           |                                                                                              |                            |
| 13   | Mon Nov 18 | Dynamical systems                                   |           |                                                                                              |                            |
|      | Wed Nov 20 | Hamiltonian Monte Carlo                             |           |                                                                                              | hw 9                       |
|      | Thu Nov 21 | lab                                                 |           |                                                                                              |                            |
| 14   | Mon Nov 25 | Special topic                                       |           |                                                                                              |                            |
|      | Wed Nov 27 | NO CLASS                                            |           |                                                                                              |                            |
|      | Thu Nov 28 | NO CLASS                                            |           |                                                                                              |                            |
| 15   | Mon Dec 02 | Special topic                                       |           |                                                                                              |                            |
|      | Wed Dec 04 | review                                              |           |                                                                                              |                            |
|      | Thu Dec 05 | lab                                                 |           |                                                                                              |                            |
:::

<!-- | Week | Date       | Topic                                          | Reading |                                    Notes                                     | Assignment                  | -->

<!-- |:-----------|:-----------|:-----------|------------|:----------:|:-----------| -->

<!-- | 1    | Mon Aug 28 | lab: welcome                                   |         |                       [💻](/slides/lab0-welcome.html)                        | [hello R](/labs/lab0.html)  | -->

<!-- |      | Tue Aug 29 | intro, history, notation                       | Ch. 2   |                                                                              | [hw 0](/hw/hw00.html)       | -->

<!-- |      | Thu Aug 31 | probability, exchangeability                   | Ch. 2   |                     [💻](/notes/lec01-probability.html)                      | [hw 1](/hw/hw01.html)       | -->

<!-- | 2    | Mon Sep 04 | NO LAB                                         |         |                                                                              |                             | -->

<!-- |      | Tue Sep 05 | single parameter estimation                    | Ch. 3   |                      [💻](/notes/lec02-estimation.html)                      |                             | -->

<!-- |      | Thu Sep 07 | Poisson model and conjugacy                    | Ch. 3   |                                                                              | [hw 2](/hw/hw02.html)       | -->

<!-- | 3    | Mon Sep 11 | lab: MLE and MAP estimator                     |         |                           [💻](/slides/lab1.html)                            |                             | -->

<!-- |      | Tue Sep 12 | reliability, exp. families                     | Ch. 3   | [💻](/notes/lec03-reliability.html), [📝](/notes/ReliabilityExpFamilies.pdf) |                             | -->

<!-- |      | Thu Sep 14 | prediction, Monte Carlo intro                  | Ch. 4   |    [💻](/notes/lec04-prediction.html), [📝](/notes/predictionOffline.pdf)    | [hw 3](/hw/hw03.html)       | -->

<!-- | 4    | Mon Sep 18 | lab: prior sensitivity and change of variables |         |                           [💻](/slides/lab2.html)                            |                             | -->

<!-- |      | Tue Sep 19 | Monte Carlo integration                        | Ch. 4   |                      [💻](/notes/lec05-MonteCarlo.html)                      |                             | -->

<!-- |      | Thu Sep 21 | the normal model                               | Ch. 5   |                     [💻](/notes/lec06-normalModel.html)                      |                             | -->

<!-- | 5    | Mon Sep 25 | practice and review                            |         |                        [💻](/slides/lab3-review.html)                        |                             | -->

<!-- |      | Tue Sep 26 | catch up / review                              |         |                                                                              |                             | -->

<!-- |      | Thu Sep 28 | [Exam I](/notes/exam-notes.html)               |         |                                                                              |                             | -->

<!-- | 6    | Mon Oct 02 | NO LAB                                         |         |                                                                              |                             | -->

<!-- |      | Tue Oct 03 | the normal model II                            | Ch. 5   |                                                                              | [hw 4](/hw/hw04.html)       | -->

<!-- |      | Thu Oct 05 | estimators                                     | Ch. 5   |     [💻](/notes/lec07-estimators.html), [📝](/notes/estimator-notes.pdf)     |                             | -->

<!-- | 7    | Mon Oct 09 | lab: predictive checks and bias                |         |                           [💻](/slides/lab4.html)                            |                             | -->

<!-- |      | Tue Oct 10 | Gibbs sampling                                 | Ch. 6   |                        [💻](/notes/lec08-gibbs.html)                         | [ec](/hw/extra-credit.html) | -->

<!-- |      | Thu Oct 12 | MCMC diagnostics                               | Ch. 6   |                   [💻](/notes/lec09-mcmc-diagnostics.html)                   | [hw 5](/hw/hw05.html)       | -->

<!-- | 8    | Mon Oct 16 | NO LAB                                         |         |                                                                              |                             | -->

<!-- |      | Tue Oct 17 | NO CLASS                                       |         |                                                                              |                             | -->

<!-- |      | Thu Oct 19 | multivariate normal (mvn)                      | Ch. 7   | [💻](/notes/lec10-mvn.html)                                                                              |                             | -->

<!-- | 9    | Mon Oct 23 | full conditional review                                               |         |                                                                              |                             | -->

<!-- |      | Tue Oct 24 | mvn parameter estimation                       | Ch. 7   |  [💻](/notes/lec11-missing-data-mvn.html), [📝](/notes/missingDataOffline.pdf)                                                                             | [hw 6](/hw/hw06.html)                        | -->

<!-- |      | Thu Oct 26 | hierarchical modeling intro                    | Ch. 8   | [💻](/notes/lec12-hierarchical-intro.html)                                                                            |                             | -->

<!-- | 10   | Mon Oct 30 | traceplots and MCMC diagnostics                                              |         | [💻](/slides/lab6-mcmc-d-practice.html)                                                                               |                             | -->

<!-- |      | Tue Oct 31 | intro to Bayesian regression                       | Ch. 9   | [💻](/notes/lec13-regression-intro.html)                                                                                |    [hw 7](/hw/hw07.html)                     | -->

<!-- |      | Thu Nov 02 | Bayesian regression II                          | Ch. 9   | [💻](/notes/lec14-BayesianRegression2.html)                                                                              |                            | -->

<!-- | 11   | Mon Nov 06 | Hierarchical modeling and Gibbs sampling practice                                              |         |                                                                              |                             | -->

<!-- |      | Tue Nov 07 | Bayesian regression III <br> *Guest lecture:* Prof. Peter Hoff                     | Ch. 9   |                                                                              | [hw 8](/hw/hw08.html)                        | -->

<!-- |      | Thu Nov 09 | NO CLASS: read [chapter summaries](/chapterSummaries.html)                                               |         |                                                                              |                             | -->

<!-- | 12   | Mon Nov 13 |  exam practice                                              |         |                [💻](/slides/lab7-exam2-prep.html)                                                                |                             | -->

<!-- |      | Tue Nov 14 | review                                         |         |                                                                              |                             | -->

<!-- |      | Thu Nov 16 |  [Exam II](/notes/exam-notes-2.pdf)                                       |         |                                                                              |                             | -->

<!-- | 13   | Mon Nov 20 | NO LAB                                         |         |                                                                              |                             | -->

<!-- |      | Tue Nov 21 |  Bayesian regression example + stan intro                                          |       |                                                                           |                             | -->

<!-- |      | Thu Nov 23 | NO CLASS                                       |         |                                                                              |                             | -->

<!-- | 14   | Mon Nov 27 |  rstanarm                                            |         |                                               [💻](/slides/lab8-rstan.html)                                 |                             | -->

<!-- |      | Tue Nov 28 | intro to Metropolis algorithm                                             |   Ch. 10      | [📝](/notes/Metropolis-intro.pdf)                                                                             | [hw 9](/hw/hw09.html)                        | -->

<!-- |      | Thu Nov 30 |  Metropolis-Hastings                                              |  Ch. 10        |  [💻](/notes/lec15-MetropolisHastings.html)                                                                         |                             | -->

<!-- | 15   | Mon Dec 04 | MCMC practice                                             |         |       [💻](/slides/lab9-MH-MCMC.html)                                                                       |                             | -->

<!-- |      | Tue Dec 05 | MCMC and HMC                                              | Ch. 10         |   [💻](/notes/lec16-MCMC-and-HMC.html)                                                                              |                             | -->

<!-- |      | Thu Dec 07 |   final review                                             |         |                                                                              |                             | -->
