---
title: "Bayesian Statistical Modeling and Data Analysis"
subtitle: "Spring 2025"
mainfont: Lato
format:
   html: 
    page-layout: full
    toc: false
---

# Schedule

```{r load-libraries, echo = F, message = F, eval = F}
library(tidyverse)
library(knitr)
```

```{r smart-date-setup, echo = F, message = F, eval = F}
# set these:
firstDay = as.Date("2025-01-08")
firstDayID = "Wed"
lastDay = as.Date("2025-4-16")
lectureDays = c("Wed", "Fri")
labDays = c("Mon")
holidays = as.Date(c("2025-01-20", "2025-03-10", "2025-03-12", "2025-03-14"))

### render table
allDates = seq(from = firstDay, to = lastDay, by = 'day')
schedule = data.frame(allDates,
                        dayOfWeek = weekdays(allDates, abbreviate=T)) %>%
  filter(dayOfWeek %in% c(lectureDays, labDays))
schedule = schedule %>%
  mutate(lab = ifelse(dayOfWeek %in% labDays, TRUE, FALSE)) %>%
  mutate(Topic = ifelse(allDates %in% holidays, "NO CLASS", "")) %>%
  mutate(Date = format(allDates, format = "%b %d")) %>%
  mutate(Date = paste(dayOfWeek, Date)) %>%
  mutate(Week = rep("", nrow(schedule))) %>%
  mutate(Reading = rep("", nrow(schedule))) %>%
  mutate(Notes = rep("", nrow(schedule))) %>%
  mutate(Assignment = rep("", nrow(schedule))) %>%
  mutate(Exam = rep("", nrow(schedule))) %>%
  mutate(Project = rep("", nrow(schedule))) %>%
  select(Week, Date, Topic, Reading, Notes, Assignment)
# to be pasted into console then pasted into markdown:
schedule %>%
  kable(format = "markdown")
```

```{=html}
<style>
    .heatMap {
        width: 100%;
        text-align: center;
    }
    .heatMap th {
        word-wrap: break-word;
        text-align: center;
    }
<!--     .heatMap tr:nth-of-type(2n+1) { -->
<!--   background: #FFF; -->
<!-- } -->
    .heatMap tr:nth-of-type(3n) { background: #D3D3D3; }
</style>
```
::: heatMap
| Week | Date       | Topic                                               | Reading |              Notes              | Assignment                 |
|:-----------|:-----------|:-----------|:----------:|:----------:|:-----------|
| 1    | Wed Jan 08 | lab: welcome                                        |         | [ðŸ’»](/slides/lab0-welcome.html) | [hello R](/labs/lab0.html) |
|      | Fri Jan 10 | intro, history, notation                            |         |                                 |                            |
| 2    | Mon Jan 13 | lab: MLE                                            |         |                                 |                            |
|      | Wed Jan 15 | probability, exchangeability                        |         |                                 |                            |
|      | Fri Jan 17 | beta-binomial model                                 |         |                                 |                            |
| 3    | Mon Jan 20 | NO CLASS                                            |         |                                 |                            |
|      | Wed Jan 22 | Poisson-gamma model, exp families                   |         |                                 |                            |
|      | Fri Jan 24 | reliability (conf. intervals, hpd, Laplace approx.) |         |                                 |                            |
| 4    | Mon Jan 27 | lab: exp. families and transformations              |         |                                 |                            |
|      | Wed Jan 29 | intro to Monte Carlo                                |         |                                 |                            |
|      | Fri Jan 31 | predictive checks and MC error                      |         |                                 |                            |
| 5    | Mon Feb 03 | lab: mixture densities                              |         |                                 |                            |
|      | Wed Feb 05 | the normal model                                    |         |                                 |                            |
|      | Fri Feb 07 | the normal model II                                 |         |                                 |                            |
| 6    | Mon Feb 10 | lab: normal data                                    |         |                                 |                            |
|      | Wed Feb 12 | estimators                                          |         |                                 |                            |
|      | Fri Feb 14 |                                                     |         |                                 |                            |
| 7    | Mon Feb 17 |                                                     |         |                                 |                            |
|      | Wed Feb 19 | Tentative Exam Date                                 |         |                                 |                            |
|      | Fri Feb 21 | Metropolis algorithm                                |         |                                 |                            |
| 8    | Mon Feb 24 |                                                     |         |                                 |                            |
|      | Wed Feb 26 | MCMC diagnostics                                    |         |                                 |                            |
|      | Fri Feb 28 | diagnostics and MH                                  |         |                                 |                            |
| 9    | Mon Mar 03 |                                                     |         |                                 |                            |
|      | Wed Mar 05 | Gibbs sampling                                      |         |                                 |                            |
|      | Fri Mar 07 | multivariate normal                                 |         |                                 |                            |
| 10   | Mon Mar 10 | NO CLASS                                            |         |                                 |                            |
|      | Wed Mar 12 | NO CLASS                                            |         |                                 |                            |
|      | Fri Mar 14 | NO CLASS                                            |         |                                 |                            |
| 11   | Mon Mar 17 |                                                     |         |                                 |                            |
|      | Wed Mar 19 | Bayesian regression I                               |         |                                 |                            |
|      | Fri Mar 21 | Bayesian regression II                              |         |                                 |                            |
| 12   | Mon Mar 24 |                                                     |         |                                 |                            |
|      | Wed Mar 26 | hierarchical modeling                               |         |                                 |                            |
|      | Fri Mar 28 | review                                              |         |                                 |                            |
| 13   | Mon Mar 31 |                                                     |         |                                 |                            |
|      | Wed Apr 02 | Tentative Exam Date                                 |         |                                 |                            |
|      | Fri Apr 04 | model averaging                                     |         |                                 |                            |
| 14   | Mon Apr 07 |                                                     |         |                                 |                            |
|      | Wed Apr 09 | Hamiltonian Monte Carlo                             |         |                                 |                            |
|      | Fri Apr 11 | Bayesian inverse problems                           |         |                                 |                            |
| 15   | Mon Apr 14 |                                                     |         |                                 |                            |
|      | Wed Apr 16 | practice for final                                  |         |                                 |                            |
:::

<!-- | Week | Date       | Topic                                               | Reading                                                |                                            Notes                                             | Assignment                 | -->

<!-- |:-----|:-----------|:----------------------------------------------------|--------------------------------------------------------|:--------------------------------------------------------------------------------------------:|:---------------------------| -->

<!-- | 1    | Mon Aug 26 | intro, history, notation                            | Ch. 2                                                  |                                                                                              | [hw 0](/hw/hw00.html)      | -->

<!-- |      | Wed Aug 28 | probability, exchangeability                        | Ch. 2                                                  |       [ðŸ—’](/notes/lec01-probability.html) [ðŸ“](/notes/Probability-Exchangeability.pdf)        | [hw 1](/hw/hw01.html)      | -->

<!-- |      | Thu Aug 29 | lab: welcome                                        |                                                        |                               [ðŸ’»](/slides/lab0-welcome.html)                                | [hello R](/labs/lab0.html) | -->

<!-- | 2    | Mon Sep 02 | NO CLASS                                            | Ch. 3                                                  |                                                                                              |                            | -->

<!-- |      | Wed Sep 04 | beta-binomial model                                 | Ch. 3                                                  |               [ðŸ—’](/notes/lec02-estimation.html) [ðŸ“](/notes/beta-binomial.pdf)               | [hw 2](/hw/hw02.html)      | -->

<!-- |      | Thu Sep 05 | lab: MLE and MAP estimator                          |                                                        |                                   [ðŸ’»](/slides/lab1.html)                                    |                            | -->

<!-- | 3    | Mon Sep 09 | Poisson-gamma model, exp families                   | Ch. 3                                                  | [ðŸ—’](notes/lec03-Poisson-gamma-exp-families.html) [ðŸ“](/notes/Poisson-gamma-exp-families.pdf) |                            | -->

<!-- |      | Wed Sep 11 | reliability (conf. intervals, hpd, Laplace approx.) | Ch. 3                                                  |         [ðŸ—’](/notes/lec04-reliability.html) [ðŸ“](/notes/reliability-handwritten.pdf)          | [hw 3](/hw/hw03.html)      | -->

<!-- |      | Thu Sep 12 | lab: exp. families and transformations              |                                                        |                             [ðŸ’»](/slides/lab-exp-families.html)                              |                            | -->

<!-- | 4    | Mon Sep 16 | intro to Monte Carlo                                | Ch. 4                                                  |         [ðŸ—’](/notes/lec05-introMonteCarlo.html) [ðŸ“](/notes/posteriorPredictive.pdf)          |                            | -->

<!-- |      | Wed Sep 18 | predictive checks and MC error                      | Ch. 4                                                  |                       [ðŸ—’](/notes/lec06-MonteCarloPredictionError.html)                       | [hw 4](/hw/hw04.html)      | -->

<!-- |      | Thu Sep 19 | lab: mixture densities                              |                                                        |                               [ðŸ’»](/slides/lab-mixtures.html)                                |                            | -->

<!-- | 5    | Mon Sep 23 | the normal model                                    | Ch. 5                                                  |             [ðŸ—’](/notes/lec07-normalModel.html)[ðŸ“](/notes/completeTheSquare.pdf)             |                            | -->

<!-- |      | Wed Sep 25 | the normal model II                                 | Ch. 5                                                  |                                 [ðŸ“](notes/normalNotes2.pdf)                                 |                            | -->

<!-- |      | Thu Sep 26 | lab: normal data                                    |                                                        |                                [ðŸ’»](/slides/lab-normal.html)                                 |                            | -->

<!-- | 6    | Mon Sep 30 | review                                              |                                                        |                                                                                              |                            | -->

<!-- |      | Wed Oct 02 | [Exam I](/notes/exam-notes.html)                    |                                                        |                                                                                              |                            | -->

<!-- |      | Thu Oct 03 | NO LAB                                              |                                                        |                                                                                              |                            | -->

<!-- | 7    | Mon Oct 07 | Metropolis algorithm                                | Ch. 10                                                 |                         [ðŸ—’](/notes/lec08-Metropolis-algorithm.html)                          |                            | -->

<!-- |      | Wed Oct 09 | MCMC diagnostics                                    | Ch. 6                                                  |                                 [ðŸ—’](/notes/lec09-MCMC.html)                                  | [hw 5](/hw/hw05.html)      | -->

<!-- |      | Thu Oct 10 | lab: Metropolis algo.                               |                                                        |                               [ðŸ’»](/slides/lab9-MH-MCMC.html)                                |                            | -->

<!-- | 8    | Mon Oct 14 | NO CLASS                                            |                                                        |                                                                                              |                            | -->

<!-- |      | Wed Oct 16 | wrap-up diagnostics and MH                          | Ch. 6, 10                                              |                             [ðŸ“](notes/ergodicMarkovChains.pdf)                              |                            | -->

<!-- |      | Thu Oct 17 | lab: MCMC and conf. bands                           |                                                        |                               [ðŸ’»](/slides/lab-conf-band.html)                               |                            | -->

<!-- | 9    | Mon Oct 21 | Gibbs sampling                                      | Ch. 6                                                  |                             [ðŸ—’](notes/lec10-gibbs-sampling.html)                             |                            | -->

<!-- |      | Wed Oct 23 | multivariate normal                                 | Ch. 7                                                  |                                  [ðŸ—’](notes/lec11-mvn.html)                                   | [hw 6](/hw/hw06.html)      | -->

<!-- |      | Thu Oct 24 | lab: MCMC diagnostics                               |                                                        |                            [ðŸ’»](slides/lab6-mcmc-d-practice.html)                            |                            | -->

<!-- | 10   | Mon Oct 28 | intro to Bayesian regression                        | Ch. 9                                                  |            [ðŸ—’](notes/lec12-regression-intro.html)[ðŸ“](notes/intro-regression.pdf)            |                            | -->

<!-- |      | Wed Oct 30 | Bayesian regression II                              | Ch. 9                                                  |                          [ðŸ—’](notes/lec13-BayesianRegression2.html)                           | [hw 7](/hw/hw07.html)      | -->

<!-- |      | Thu Oct 31 | lab: rstanarm                                       |                                                        |                                 [ðŸ’»](slides/lab8-rstan.html)                                 |                            | -->

<!-- | 11   | Mon Nov 04 | estimators                                          | Ch. 5 sec. 4                                           |              [ðŸ—’](notes/lec14-estimators.html)[ðŸ“](notes/estimator-notes-24.pdf)              |                            | -->

<!-- |      | Wed Nov 06 | hierarchical modeling                               | Ch. 8                                                  |          [ðŸ—’](notes/lec15-hierarchical-intro.html)[ðŸ“](notes/hierarchical-notes.pdf)          | [hw 8](/hw/hw08.html)      | -->

<!-- |      | Thu Nov 07 | lab: estimators                                     |                                                        |                               [ðŸ’»](slides/estimator-lab.html)                                |                            | -->

<!-- | 12   | Mon Nov 11 | review                                              |                                                        |                                                                                              |                            | -->

<!-- |      | Wed Nov 13 | [Exam II](/notes/exam-notes.html)                   |                                                        |                                                                                              |                            | -->

<!-- |      | Thu Nov 14 | NO LAB                                              |                                                        |                                                                                              |                            | -->

<!-- | 13   | Mon Nov 18 | priors                                              |                                                        |                                    [ðŸ—’](notes/priors.html)                                    |                            | -->

<!-- |      | Wed Nov 20 | Bayesian inverse problems                           | [ðŸ“–](https://www.pnas.org/doi/10.1073/pnas.2409983121) |                                 [ðŸ—’](notes/waterOnMars.html)                                  | [hw 9](/hw/hw09.html)      | -->

<!-- |      | Thu Nov 21 | lab: Bayesian inverse practice                      |                                                        |                          [ðŸ’»](/slides/BayesianInverseProblem.html)                           |                            | -->

<!-- | 14   | Mon Nov 25 | Hamiltonian Monte Carlo                             | [ðŸ“–](/readings/readHMC.html)                           |                            [ðŸ—’](notes/HamiltonianMonteCarlo.html)                             |                            | -->

<!-- |      | Wed Nov 27 | NO CLASS                                            |                                                        |                                                                                              |                            | -->

<!-- |      | Thu Nov 28 | NO CLASS                                            |                                                        |                                                                                              |                            | -->

<!-- | 15   | Mon Dec 02 | model averaging                                     | Ch. 9 sec. 3                                           |                            [ðŸ—’](notes/BayesianModelAveraging.html)                            |                            | -->

<!-- |      | Wed Dec 04 | practice for final                                  |                                                        |                                [ðŸ—’](notes/practiceFinal.html)                                 |                            | -->

<!-- |      | Thu Dec 05 | lab: review                                         |                                                        |                                                                                              |                            | -->

<!-- ::: -->
