---
title: "Exponential families & variable transformation"
format: 
    revealjs:
      mainfont: Lato
      smaller: true
---


# Practice exercises

## Ex 1: exponential family


Let $p(y|\theta) = \theta^{y}(1-\theta)^{1-y}$

a. Identify the transform $\phi = f(\theta)$ such that $p(y | \phi)$ can be written as $h(y) c(\phi) e^{\phi t(y)}$. Identify the sufficient statistic $t(y)$, as well as $c(\phi)$.
b. Write down the prior $p(\phi|n_0, t_0)$.



## Ex 2: variable transform

Confirm that $f(\theta)$ from the previous exercise is monotonic and invertible. Next, using the one-line formula: $p_\theta(\theta) = p_{\phi}(f(\theta))|\frac{d\phi}{d\theta}|$, show that the prior, $p(\theta | n_0, t_0)$ is a beta density.

To very the above numerically, sample from $p(\theta | n_0, t_0)$ using the code below, for some $n_0, t_0$ of your choosing.

```
theta = rbeta(n = 10000, shape1 = n_0 * t_0, shape2 = n_0*(1-t_0)
```

Next, transform each sample of the object `theta` to `phi` using $f(\theta)$ as defined in exercise 1.

Finally, plot a density plot of samples `phi` and, on the same plot, add $p(\phi | n_0, t_0)$ from exercise 1, part (c).

## Ex 3: another variable transform

Let $X \sim \text{Unif}(5, 10)$ and let $Y = X^2$

Notice that even though $X^2$ is not a monotonic function *everywhere*, it *is* a monotonic function **over the support of X**. 

Exercise: use the change of variables formula to derive $p(y)$. Confirm via simulation, as in exercise 2.

# Solutions

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
```


## Ex 1: solution

#### a

$$
\phi = log(\frac{\theta}{1-\theta})
$$

Therefore,

$$
p(y|\phi) = e^{\phi y} \left(\frac{1}{1 + e^{\phi}} \right)
$$

and

$$
\begin{aligned}
t(y) &= y\\
c(\phi) &= \left(\frac{1}{1 + e^{\phi}} \right)
\end{aligned}
$$

#### b

$$
p(\phi | n_0, t_0) = \left({1 + e^{\phi}} \right)^{-n_0} e^{n_0 t_0 \phi}
$$

## Ex 2: solution

Monotonic if $\theta_1 \geq \theta_2$ implies $f(\theta_1) \geq f(\theta_2)$ over the support $\theta_1, \theta_2 \in [0,1]$.

Let $\theta_1 > \theta_2$,  let's check:

$$
\begin{aligned}
\log \left(\frac{\theta_1}{1-\theta_1}\right) &> \log \left(\frac{\theta_2}{1-\theta_2}\right)\\
\theta_1 (1-\theta_2) &> \theta_2(1-\theta_1)\\
\theta_1 > \theta_2
\end{aligned}
$$
It is monotonic!

Furthermore, we found it was invertible in ex 1: $\theta = e^{\phi}/(1 + e^{\phi})$.

One-line formula: $p_\theta(\theta) = p_{\phi}(f(\theta))|\frac{d\phi}{d\theta}|$:

$$
\begin{aligned}
p(\theta) &= (1-\theta)^{n_0} (\theta/(1-\theta))^{n_0 t_0} \left|\frac{d\phi}{d\theta} \right|\\
&= \theta^{n_0 t_0 -1} (1-\theta)^{n_0(1 -t_0) - 1}
\end{aligned}
$$
So $\theta \sim \text{beta}(n_0 t_0, n_0(1-t_0))$.

Remember that $a$ and $b$ of a beta distribution must both be positive! This constrains what $n_0$ and $t_0$ can be.

## Ex 2: solution (pt 2)

::: panel-tabset

## plot

```{r}
#| warning: false
#| message: false
#| echo: false

n_0 = 5
t_0 = .5
theta = rbeta(n = 100000, shape1 = n_0 * t_0, shape2 = n_0 * (1 - t_0))
phi = log((theta / (1-theta)))

df = data.frame(phi)

fphi = function(phi, n_0, t_0) {
  ( (1 + exp(phi))^(-n_0) ) * exp(n_0 * t_0 * phi) / 
    beta(n_0 * t_0, n_0 * (1 - t_0))
}

df %>%
  ggplot(aes(x = phi)) + 
  stat_function(fun = fphi, args = list(n_0 = n_0, t_0 = t_0)) +
  geom_histogram(aes(x = phi, y = ..density..),
                 fill = 'steelblue', alpha = 0.5)
```



## code
```{r}
#| eval: false

n_0 = 5
t_0 = .5
theta = rbeta(n = 100000, shape1 = n_0 * t_0, shape2 = n_0 * (1 - t_0))
phi = log((theta / (1-theta)))

df = data.frame(phi)

fphi = function(phi, n_0, t_0) {
  ( (1 + exp(phi))^(-n_0) ) * exp(n_0 * t_0 * phi) / 
    beta(n_0 * t_0, n_0 * (1 - t_0))
}

df %>%
  ggplot(aes(x = phi)) + 
  stat_function(fun = fphi, args = list(n_0 = n_0, t_0 = t_0)) +
  geom_histogram(aes(x = phi, y = ..density..),
                 fill = 'steelblue', alpha = 0.5)
```
:::

The normalizing constant can be obtained by transforming from $p(\theta)$ (which is beta) back to $p(\phi)$.


## Ex 3: solution

::: panel-tabset

## plot

```{r}
#| warning: false
#| echo: false

x = runif(100000, 5, 10)
y = x^2

df = data.frame(y)

f = function(y) {
  return(.1/sqrt(y))
}

df %>%
  ggplot(aes(x = y)) + 
  stat_function(fun = f) +
  geom_histogram(aes(x = y, y = ..density..),
                 fill = 'steelblue', alpha = 0.5)
```


## code 
```{r}
#| warning: false
#| eval: false

x = runif(100000, 5, 10)
y = x^2

df = data.frame(y)

f = function(y) {
  return(.1/sqrt(y))
}

df %>%
  ggplot(aes(x = y)) + 
  stat_function(fun = f) +
  geom_histogram(aes(x = y, y = ..density..),
                 fill = 'steelblue', alpha = 0.5)
```


:::