---
title: "Sensitivity to the prior and change of variables"
format: 
    revealjs:
      mainfont: Lato
      smaller: true
---

# Sensitivity to the prior

```{r}
#| echo: false
#| warning: false
library(tidyverse)
```

## Practice exercise

A cancer laboratory is estimating the rate of tumorigenesis in two strains of mice, $A$ and $B$. They have tumor count data for 10 mice in strain $A$ and 13 mice in strain $B$,

```{r}
yA = c(12, 9, 12, 14, 13, 13, 15, 8, 15, 6)
yB = c(11, 11, 10, 9, 9, 8, 7, 10, 6, 8, 8, 9, 7)
```

Assume

$$
\begin{aligned}
Y_A &\sim \text{Poisson}(\theta_A)\\
Y_B &\sim \text{Poisson}(\theta_B).
\end{aligned}
$$

::: panel-tabset

## Exercise 1

Let

$$
\begin{aligned}
\theta_A &\sim \text{gamma}(120, 10)\\
\theta_B &\sim \text{gamma}(12, 1).
\end{aligned}
$$

- Compute $p(\theta_B < \theta_A ~|~ \vec{y}_A, \vec{y}_B)$ via Monte Carlo sampling.

## Exercise 2

Let 

$$
\begin{aligned}
\theta_A &\sim \text{gamma}(120, 10)\\
\theta_B &\sim \text{gamma}(12\cdot n_0, n_0).
\end{aligned}
$$

- For a range of values of $n_0$, obtain $p(\theta_B < \theta_A ~|~ \vec{y}_A, \vec{y}_B)$. 

- Describe how sensitive conclusions about the event $\{ \theta_B < \theta_A\}$ are to the prior distribution on $\theta_B$.

:::

# Change of variables

## Practice exercise

- Let $X \sim \text{Unif}(5, 10)$
- Let $Y = X^2$

Notice that even though $X^2$ is not a monotonic function *everywhere*, it *is* a monotonic function **over the support of X**. 

Exercise: use the change of variables formula to derive $p(y)$. Confirm with Monte Carlo simulation.



```{r}
#| code-fold: true
#| code-summary: "Show solution"
#| warning: false
#| eval: false

library(tidyverse)

x = runif(100000, 5, 10)
y = x^2

df = data.frame(y)

f = function(y) {
  return(.1/sqrt(y))
}

df %>%
  ggplot(aes(x = y)) + 
  stat_function(fun = f) +
  geom_histogram(aes(x = y, y = ..density..),
                 fill = 'steelblue', alpha = 0.5)
```

