---
title: "Extra practice"
format: 
    revealjs:
      mainfont: Lato
      smaller: true
---

# Predictive checks

## Exercise

A data scientist at a small subscriber-based tech company models the number of new subscribers in a day as $Y|\theta \sim \text{Poisson}(\theta)$ with prior $\theta \sim \text{gamma}(a,b)$. A priori, the data scientist believes that there are on average 20 signups per day and 90% of the time there are between approximately 3 and 50 signups on a given day.

#### a

Find suitable $a$ and $b$ that satisfy the data scientist's prior beliefs.

Verify how well your prior aligns with this belief using Monte Carlo sampling to generate the prior predictive distribution, $p(\tilde{y}) = \int p(\tilde{y}, \theta)d\theta$.

```{r}
#| eval: false
#| echo: false
set.seed(360)
a = 2; b = .1

N = 10000
theta.prior = rgamma(N, a, b)
y.prior = rpois(N, theta.prior)
quantile(y.prior, c(0.05, 0.95))
mean(y.prior) # equivalently, a/b = 20
```

#### b

After one month the data scientist observes the following daily subscriber counts:

```{r}
y = c(10, 21, 19, 16, 20, 18, 35, 16, 23, 26, 20, 21, 23, 19, 18, 20, 23, 18, 21, 16, 15, 15, 20, 22, 19, 25)
```

The data scientist is fundamentally interested in the variance of subscriber counts per day. Is the Poisson model appropriate for this data?

Report $p(\tilde{S}^2 > s^2_{obs} | y_1,\ldots y_n)$ where $\tilde{S}^2$ is the posterior predictive sample variance and $s^2_{obs}$ is the observed sample variance ($s^2_{obs} = 21.3$). To generate samples under the posterior predictive distribution, use the prior from part (a).

```{r}
#| echo: false
#| eval: false
set.seed(360)

# posterior predictive distribution
ytotal = sum(y)
a = 2 ; b = .1; n = length(y)
s2 = var(y) # test statistic

t.mc = NULL
for (i in 1:10000) {
  theta1 = rgamma(1, ytotal + a, b + n) # draw 1 theta from posterior
  y.mc = rpois(n, theta1) # draw y from post pred of sample size n = 30
  t.mc = c(t.mc, var(y.mc))# compute t
}

tailProb = mean(t.mc > s2)

tailProb

# tail probability of 0.363
# the Poisson model is probably suitable for making inference about the variance here
```


# Estimator bias

## Exercise

Let $Y_1,\ldots Y_n$ be iid random variables with expectation $\theta$ and variance $\sigma^2$.

Show that $\frac{1}{n} \sum_{i = 1}^n (Y_i -\bar{Y})^2$ is a biased estimator of $\sigma^2$.