---
title: "MCMC diagnostics practice"
format: 
    revealjs:
      mainfont: Lato
      smaller: true
---

# Exercises

## Exercise 1

Let $p(\theta_1, \theta_2 | \mathbf{y})$ be our **target distribution**, i.e. the distribution we are interested in sampling.

We construct a Gibbs sampler and look at the trace plots of $\theta_1$ and $\theta_2$, produced below. Chat with your neighbor, describe what you observe. Has the chain **converged** for each parameter? How well is the sampler **mixing**? Do you think the parameters are correlated or uncorrelated? 

Write down a description of the plots below, in 2 paragraphs or less, for a reader who has **not taken this class**. You may assume the reader has taken a course on probability (e.g. similar to STA230/240).


```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(latex2exp)
library(patchwork)
theta1 = c(rnorm(750, mean = 10, sd = 1), rnorm(250, mean = 50, sd = 4))
theta2 = c(rnorm(750, mean = 45, sd = 2), rnorm(250, mean = 53, sd = 1))

df = data.frame(theta1, theta2)
p1 = df %>%
ggplot(aes(x = 1:1000)) +
  geom_line(aes(y = theta1)) +
  theme_bw() +
  labs(y = TeX("$\\theta_1$"), x = "iteration")

p2 = df %>%
ggplot(aes(x = 1:1000)) +
  geom_line(aes(y = theta2)) +
  theme_bw() +
  labs(y = TeX("$\\theta_2$"), x = "iteration")

p1 + p2
```

## Exercise 2

Based on the first 1000 iterations of your Gibbs sampler shown on the previous slide, which of the following joint densities is the most plausible for $\theta_1, \theta_2 | \mathbf{y}$? Why? Hint: it may help to think about where your sampler starts and imagine a particle moving through space according to conditional updates.

```{r}
#| echo: false
#| warning: false

theta1_true = c(rnorm(3000, mean = 10, sd = 1), rnorm(1000, mean = 50, sd = 4))
theta2_true = c(rnorm(3000, mean = 45, sd = 2), rnorm(1000, mean = 53, sd = 1))

theta1_base =  c(rnorm(2000, mean = 10, sd = 1), rnorm(1000, mean = 50, sd = 4),
                 rnorm(1000, mean = 80, sd = 3))

theta2_base = c(rnorm(2000, mean = 45, sd = 2), rnorm(1000, mean = 53, sd = 1),
                rnorm(1000, mean = 35, sd = 2))
  

df_base = data.frame(theta1 = theta1_base, theta2 = theta2_base)
df_true = data.frame(theta1 = theta1_true, theta2 = theta2_true)

p1 = df_true %>%
  mutate(theta1 = theta1 / 10,
         theta2 = theta2 / 10) %>%
  ggplot(aes(x = theta1, y = theta2)) + 
  geom_bin2d(bins = 25) +
  theme_bw() + 
  labs(y = TeX("$\\theta_2$"), 
       x = TeX("$\\theta_1$"))

df2 = df_base

df2$theta1 = rnorm(4000, mean = 50, sd = 4)

p2 = df2 %>%
  ggplot(aes(x = theta1, y = theta2)) +
  geom_bin2d(bins = 25) +
  theme_bw() +
  labs(y = TeX("$\\theta_2$"),
       x = TeX("$\\theta_1$"))

df3 = df_base
df3$theta1 = rnorm(4000, mean = 35, sd = 15)
df3$theta2 = rnorm(4000, mean = 48, sd = 10)


p3 = df3 %>%
  ggplot(aes(x = theta1, y = theta2)) +
  geom_bin2d(bins = 25) +
  theme_bw() +
  labs(y = TeX("$\\theta_2$"),
       x = TeX("$\\theta_1$"))

df4 = df_base

p4 = df4 %>%
  ggplot(aes(x = theta1, y = theta2)) +
  geom_bin2d(bins = 25) +
  theme_bw() +
  labs(y = TeX("$\\theta_2$"),
       x = TeX("$\\theta_1$"))


(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "Joint densities", tag_levels = c("A"))
```


