---
title: "Homework 4"
subtitle: "Due Friday September 27 at 5:00pm"
mainfont: Lato
format: 
  html:
    toc: true
---


## Exercise 1

Let $\theta$ be the rate of mutation for a certain cluster of cancer cells. Biologists encode their prior uncertainty about $\theta$ with the following density:

$$
p(\theta) = \frac{4}{10}e^{-4\theta} + \frac{9}{10 \Gamma(8)} \theta^7 e^{-\theta}
$$
a. Make a plot of this prior density and explain what it means, in context, to the biologists.

b. Let $Y_i$ be the number of mutations produced by the $i$th cell, such that

$$
Y_i | \theta \sim Poisson(\theta)
$$

Write out the posterior distribution of $\theta$ given $y_1,\ldots y_n$ (up to a proportionality constant) and simplify as much as possible. Hint: Be careful when writing your proportionality statement!

c. The posterior is a mixture (weighted average) of two distributions that you know. Identify these two distributions, including their parameters.

d. Assume you obtain mutation data from two cells, $y_1 = 3, y_2 = 1$. Compute the posterior exactly (i.e. find the appropriate integration constant) and plot the posterior density. 

e. In part (c) you identified the posterior as a mixture (weighted average) of two distributions. Given the data in part (d), compute the weights of each density in the mixture.

## Exercise 2

A group of scientists have mutation data, 

```{r}
y = c(0,3,0,1,5,2,0,4,1,1)
```

and are interested in assessing how well the Poisson model from exercise 1 fits their data. Using the data generative model and prior from exercise 1, generate posterior predictive datasets $y^{(1)}, y^{(2)}, \ldots y^{(S)}$, where each data set $y^{(s)}$ is a vector of length 10 whose entries are sampled from the Poisson distribution with parameters $\theta^{(s)}$. Each $\theta^{(s)}$ itself is a sample from the posterior $p(\theta | y_1, \ldots y_{10})$. For each $s$, let $t(s)$ be the sample average of the 10 values of $y^{(s)}$ divided by the sample standard deviation of $y^{(s)}$. Make a histogram of $t(s)$ and compare to the observed value of this statistic. Based on this statistic, assess the fit of the Poisson model for these data.


## Exercise 3

Chimowitz et al. (2011) [https://doi.org/10.1056/NEJMoa1105335](https://doi.org/10.1056/NEJMoa1105335) investigate if stents are effective treatment to manage strokes in patients with atherosclerotic intracranial arterial stenosis. You can load the data using the code below. Data sourced from `openintro` package.

```{r}
#| eval: false
#| message: false
stents = readr::read_csv("https://sta360-fa24.github.io/data/stent365.csv")
```

Each row of the data set is an individual patient. The `group` column indicates whether the patient was treated with a stent or not. The `outcome` column reports whether the patient had a stroke or not within a year.

a. Write down a data generative model (possibly as two separate densities) for this data.

b. Write down your prior beliefs about unknown parameters in your model above using a conjugate density. Choose parameters for the priors. Explain your choices.

c. Report (using Monte Carlo sampling or otherwise) the posterior mean of the [relative risk](https://en.wikipedia.org/wiki/Relative_risk), i.e. the posterior mean of the probability of stroke in the treatment group versus in the control, (think $E~[\frac{\theta_t}{\theta_c}~|~\text{data}]$). Additionally, include a 95% posterior confidence interval for the relative risk.

d. Plot the posterior of the relative risk from part c. Do you believe the treatment is effective?