---
title: "Homework 6"
subtitle: "Due Wednesday November 1st at 5:00pm"
mainfont: Lato
format: 
  html:
    toc: true
---

## Exercise 1

6.3 from Hoff. You can simulate from a constrained normal distribution with mean `mean` and standard deviation `sd`, constrained to lie in the interval $(a,b)$ using the following function:

```{r}
rcnorm<-function(n, mean=0, sd=1, a=-Inf, b=Inf){
  u = runif(n, pnorm((a - mean) / sd), pnorm((b - mean) / sd))
  mean + (sd * qnorm(u))
}
```

Note that you can use this function to simulate a vector of constrained normal random variables, each with a potentially different mean, standard deviation, and constraints.

To load the data for this exercise, run the code below

```{r}
#| eval: false
divorce = readr::read_csv("https://sta360-fa23.github.io/data/divorce.csv")
```


## Exercise 2

Show that if $W \sim \text{Wishart}(m, S)$ then $E[W] = mS$.

## Exercise 3

Suppose $Y$ is a random normal vector $Y \sim N_p(\theta, \Sigma)$. Let $Y_A$ be the first $p_1$ elements of $Y$ and $Y_B$ be the last $p_2 = p - p_1$ elements, so that $Y = (Y_A, Y_B)$. Similarly, write $\theta = (\theta_A, \theta_B)$. Finally, let 

$$
\Sigma^{-} \equiv \Psi = \left[ {\begin{array}{cc}
   \Psi_{AA} & \Psi_{AB} \\
   \Psi_{BA} & \Psi_{BB} \\
  \end{array} } \right]
$$
and note that $\Psi_{AB} = \Psi_{BA}^T$. Find the conditional distribution of $Y_B$ given $Y_A$ in terms of $\theta_A$, $\theta_B$ and components of $\Psi$. Try to interpret how $E[Y_B|Y_A]$ differs from $E[Y_B]$ and how $V[Y_B|Y_A]$ differs from $V[Y_B]$.

### Identities for exercise 3

Some of the following identities will be helpful for *interpretation*.

Let 

$$
\Sigma = \left[ {\begin{array}{cc}
   \Sigma_{AA} & \Sigma_{AB} \\
   \Sigma_{BA} & \Sigma_{BB} \\
  \end{array} } \right]
$$

and

$$
\Psi = \left[ {\begin{array}{cc}
   \Psi_{AA} & \Psi_{AB} \\
   \Psi_{BA} & \Psi_{BB} \\
  \end{array} } \right].
$$

Then 

$$
\begin{aligned}
\Psi_{AA}^- &= \Sigma_{AA} - \Sigma_{AB} \Sigma_{BB}^- \Sigma_{BA}\\
\Psi_{BB}^- &= \Sigma_{BB} - \Sigma_{BA} \Sigma_{AA}^- \Sigma_{AB}\\
\Psi_{AB} &= -\Psi_{AA} \Sigma_{AB} \Sigma_{BB}^-\\
\Psi_{BA} &= -\Psi_{BB} \Sigma_{BA} \Sigma_{AA}^-,
\end{aligned}
$$

and note that $\Sigma_{AB} = \Sigma_{BA}^T$ and $\Psi_{AB} = \Psi_{BA}^T$.