---
title: "Parameter Estimation in Dynamical Systems"
author: "Dr. Alexander Fisher"
# mainfont: Lato
format: 
  html:
    toc: true
---

```{r}
#| label: load-packages
#| warning: false
#| message: false
library(tidyverse)
```

## Example: Hares-Lynx population dynamics

A Canadian business, *Hudson's Bay company* is North America's longest continually operating company (founded in 1670). During the 19th and 20th century, the company bought animal pelts of both snowshoe hares and lynx from trappers. Lynx are the most predominant natural predator of snowshoe hares. The counts of pelts provide an approximation to the population size of each animal. Data sourced from [here](https://jmahaffy.sdsu.edu/courses/f00/math122/labs/labj/q3v1.htm).

```{r}
#| label: load-data
#| warning: false
#| message: false

hares_lynx = read_csv("../data/hares-lynx.csv")
glimpse(hares_lynx)
```

-   `year` is the year of record
-   `hares` is the number of hare pelts (thousands)
-   `lynx` is the number of lynx pelts (thousands)

### Dynamical System

The following system is an example known as "Lotka-Volterra" dynamics which describe an interactive model of the population of predator and prey.

|              |                                |
|--------------|--------------------------------|
| $H(t)$       | Number of hares at time t      |
| $H(0) = H_0$ | Number of hares at time 0      |
| $L(t)$       | Number of lynx at time t       |
| $L(0) = L_0$ | Number of lynx at time 0       |
| $a_1$        | Per capita birth rate of hares |
| $b_1$        | Per capita death rate of lynx  |
| $a_2$        | Rate at which lynx eat hares   |
| $b_2$        | Rate at which prey affects predator growth                               |


Some assumptions: 

1. The prey population finds ample food at all times.
2. The food supply of the predator population depends entirely on the size of the prey population.
3. The rate of change of population is proportional to its size.
4. During the process, the environment does not change in favor of one species, and genetic adaptation is inconsequential.
5. Predators have limitless appetite.
6. Both populations can be described by a single variable. This amounts to assuming that the populations do not have a spatial or age distribution that contributes to the dynamics.

$$
\begin{aligned}
\frac{dH(t)}{dt} &= a_1 H(t) - a_2 H(t)L(t)\\
\frac{dL(t)}{dt} &= - b_1 L(t) + b_2 H(t)L(t)
\end{aligned}
$$


### Statistical model

One more assumption: we will assume multivariate normal noise in our observations $H(t), L(t)$.

$$
  \begin{align}
   \begin{bmatrix}
           H(t) \\
           L(t)
         \end{bmatrix}
         &\sim
         MVN \left(
         \begin{bmatrix}
           \tilde{H}(t) \\
           \tilde{L}(t)
         \end{bmatrix}, 
         \Sigma
          \right)
  \end{align}
$$

where $\tilde{H}(t), \tilde{L}(t)$ is the solution of the system of differential equations and is a function of $H(0), L(0), a_1, a_2, b_1, b_2$ as well as $t$.

$\Sigma$ is a $2 \times 2$ matrix 

We place priors on the unknowns

$$
\begin{aligned}
a_1 &\sim N()\\
a_2 &\sim \text{log-normal}\\
b_1 &\sim N()\\
b_2 &\sim \text{gamma}()
\end{aligned}
$$

### Parameter estimation

```{r}

```

