---
title: "Intro to Monte Carlo"
author: "Dr. Alexander Fisher"
mainfont: Lato
format: 
  html:
    toc: true
---

Load packages: 

```{r}
#| warning: false
library(tidyverse)
library(latex2exp)
```

## Monte Carlo motivation

[General social survey](https://gss.norc.org/) from the 90s gathered data on the number of children to women of two categories: those with and without a bachelor's degree. 

Setup:

- $Y_{i1}$: number of children of $i$th woman in group 1 (no bachelor's)
- $Y_{i2}$: number of children of $i$th woman in group 2 (bachelor's)

Model: 

- $Y_{11}, \ldots, Y_{n_1 1} | \theta_1 \overset{\mathrm{iid}}{\sim} \text{Poisson}(\theta_1)$
- $Y_{12} \ldots, Y_{n_2 2} | \theta_2 \overset{\mathrm{iid}}{\sim} \text{Poisson}(\theta_2)$

Prior:

- $\theta_1 \sim \text{gamma}(2, 1)$
- $\theta_2 \sim \text{gamma}(2, 1)$

Data: 

- $n_1 = 111$, $\bar{y_1} = 1.95$, $\sum y_{i 1} = 217$
- $n_2 = 44$, $\bar{y_1} = 1.5$, $\sum y_{i 1} = 66$

Posterior:

- $\theta_1 | \vec{y_1} \sim \text{gamma}(219, 112)$
- $\theta_2 | \vec{y_2} \sim \text{gamma}(68, 45)$


We already know how to compute 

- posterior mean: $E~\theta | y = \alpha / \beta$ (*shape*, *rate* parameterization)
- posterior density (`dgamma`)
- posterior quantiles and confidence intervals (`qgamma`)

What about posterior distribution of $|\theta_1 - \theta_2|$, $\theta_1 / \theta_2$, $\text{max} \{\theta_1, \theta_2 \}$?

What about the probability a woman with a bachelor's has more children than a woman without a bachelors? $p(\tilde{y}_1  < \tilde{y}_2 | \vec{y_1}, \vec{y_2})$?

## Monte Carlo integration

- approximates an integral by a stochastic average
- shines when other methods of integration are impossible (e.g. high dimensional integration)
- works because of law of large numbers:  for a random variable $\theta$, the sample mean $\bar{\theta}_N$ converges to the true mean $\mu$ as the number of samples $N$ tends to infinity.

The key idea is: we obtain independent samples from the posterior, 

$$
\theta^{(1)}, \ldots \theta^{(N)} \overset{\mathrm{iid}}{\sim} p(\theta |\vec{y})
$$
then the empirical distribution of the samples approximates the posterior (approximation improves as $N$ increases).

Recall

$$
E~g(\theta)|y = \int_\mathcal{\theta} g(\theta) p(\theta | y)d\theta \approx \frac{1}{N} \sum_{i = 1}^N g(\theta^{(i)}).
$$

The law of large numbers says that if our samples $\theta^{(i)}$ are independent, $\frac{1}{N} \sum_{i = 1}^N g(\theta^{(i)})$ to $E~\theta|y$.

:::callout-note
Integrals are expectations, **and** expectations are integrals.
:::

## Examples

- $\theta_1 | \vec{y_1} \sim \text{gamma}(219, 112)$
- $\theta_2 | \vec{y_2} \sim \text{gamma}(68, 45)$

#### (1) proof of concept: the mean

```{r}
set.seed(123)
N = 5000
rgamma(N, shape = 219, rate = 112) %>%
  mean()
```

Pretty close to the true mean, `r 219/112`. 

#### (2) posterior of $|\theta_1 - \theta_2|$

```{r}
set.seed(123)
theta1 = rgamma(N, shape = 219, rate = 112)
theta2 = rgamma(N, shape = 68, rate = 45)

df = data.frame(diff = abs(theta1 - theta2))

df %>%
  ggplot(aes(x = diff)) + 
  geom_density() +
  theme_bw() +
  labs(x = TeX("$|\\theta_1 - \\theta_2|$"),
       y = TeX("$p(|\\theta_1 - \\theta_2 || {y}_1, {y}_2)$"))
```

#### (3) $p(|\theta_1 - \theta_2|> .5)$

```{r}
mean(df$diff > .5)
```


::: panel-tabset
## Exercise

::: callout-important
## Exercise
- $\theta  \sim \text{uniform}(0, 2)$
- Let $\phi = \log \theta$
- Visualize $p(\phi)$ using Monte Carlo simulation, then show using the change of variables formula and plotting the closed form of the density.
:::

## Full solution

```{r}
#| warning: false

# sample from p(theta)
theta = runif(10000, 0, 2)

# define transform function
f = function(x) {
  return(0.5 *exp(x))
}

# create a df for each plot
df = data.frame(phi = -7:0)
df2 = data.frame(phiSamples = log(theta))

# make plots
df %>%
  ggplot(aes(x = phi)) +
  stat_function(fun = f, col = 'red', alpha = 0.5) +
  geom_histogram(data = df2, aes(x = phiSamples,
                                 y = ..density..),
               fill = 'steelblue', alpha = 0.5)

```

## Quick Monte Carlo

```{r}
# Just making the Monte Carlo part of the plot 
# in 3 lines
theta = runif(10000, 0, 2)
phi = log(theta)
hist(phi)
```
:::


::: panel-tabset

## Exercise 

::: callout-important
## Exercise

What is $Pr(\tilde{y}_1  < \tilde{y}_2 | \vec{y_1}, \vec{y_2})$?
:::

## Solution
```{r}
theta1 = rgamma(N, shape = 219, rate = 112)
theta2 = rgamma(N, shape = 68, rate = 45)

y1tilde = rpois(N, theta1)
y2tilde = rpois(N, theta2)

# y1: no. children to parent w/ no bachelors
# y2: no. children to parent w/ bachelors
mean(y1tilde < y2tilde)
```

:::
