---
title: "Quiz 5"
---

To receive credit for this assignment, write your full name at the top of your paper.

### Exercise 1

Let $\theta$ be a positive random variable. In the first 100 states of a Markov chain we sample $\theta \in (0, 5)$, then for states 100 through 10000 we sample $\theta$ values $\in$ $(5, \infty)$. This Markov chain is not ____. (Choose one of the following to fill in the blank: *aperiodic*, *irreducible*, *recurrent*).

### Exercise 2

Write down the two distributions you need to be able to sample from to write a Gibbs sampler for the following target: $p(\beta_1, \beta_2 | y_1, \ldots, y_n)$.  

### Exercise 3

Which traceplot reveals a Markov chain that hast not reached stationarity? 

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 2
set.seed(360)
library(tidyverse)
library(patchwork)
S = 500
THETA = NULL
BETA = NULL
theta = 0
for(i in 1:S) {
  THETA = c(THETA, theta)
  theta = rnorm(1, theta + .01, .1)
  BETA = c(BETA, rnorm(1, 5, 1))
}
df = data.frame(THETA, BETA)
p1 = df %>%
  ggplot(aes(x = 1:S)) +
  geom_line(aes(y = THETA)) +
  labs(x = "iteration", y = "theta") +
  theme_bw() +
  ggtitle('A')

p2 = df %>%
  ggplot(aes(x = 1:S)) +
  geom_line(aes(y = BETA)) +
  theme_bw() + 
  labs(x = "iteration", y = "theta") +
  ggtitle('B')


p1 + p2
```


<!-- ### Exercise 1 -->

<!-- $$ -->
<!-- Y | \theta, \sigma^2 \sim N(\theta, \sigma^2)  -->
<!-- $$ -->

<!-- What is a conjugate class of priors for $\theta | \sigma^2$? -->


<!-- ### Exercise 2 -->

<!-- `TRUE` or `FALSE` -->

<!-- An estimator is a random variable. -->

<!-- ### Exercise 3 -->

<!-- Let $\theta_0$ be the true value of $\theta$. -->

<!-- If $E[\hat{\theta}| \theta = \theta_0] = \theta_0$, we say $\hat{\theta}$ is a ____ estimator of $\theta$. -->


```{r}
#| echo: false
library(countdown)
countdown(minutes = 3, font_size = "4em",
          color_running_background = "#2774AE")
```

