---
title: "Quiz 3"
---

To receive credit for this assignment, write your full name at the top of your paper.

### Exercise 1

Write "posterior" or "prior" in the blank below:

$\int p(\tilde{y}|\theta) p(\theta | y_1,\ldots y_n)d\theta$ is a ___ predictive distribution.

### Exercise 2

Expand $p(\theta | y)$ using Bayes' rule (include the normalization constant).

### Exercise 3

The vertical line on the following plot is best described as the posterior ____. 

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 2
library(tidyverse) 
library(latex2exp)
data.frame(x = 0:6) %>%
  ggplot(aes(x = x)) +
  stat_function(fun = dgamma, args = list(2, 2)) +
  theme_bw() +
  geom_vline(xintercept = 1/2, col = "red") +
  labs(x = TeX("$\\theta$"), 
       y = TeX("$p(\\theta | y_1,\\ldots y_n)$")) +
  theme(text = element_text(size = 15))
```



<!-- `TRUE` or `FALSE` -->

<!-- Monte Carlo integration error scales $\mathcal{O}(\frac{1}{\sqrt{N}})$ where $N$ is the number of independent samples from the posterior. -->


```{r}
#| echo: false
library(countdown)
countdown(minutes = 3, font_size = "4em",
          color_running_background = "#2774AE")
```

